---
title: "reproducible-res-project2.Rmd"
output: html_document
---

# **Title:** 
### Analysis of severe weather effects on US public health and economy

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(ggplot2)
library(reshape2)
```

## **Synopsis:**
######Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
###### This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

###### This analysis will summarize effects of extereme wether conditions.


## **Data Processing**

Download bz2 file and load table into memory

```{r}
readData <- function()
{
  dir.create("temp",showWarnings = FALSE)
  dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  if(!file.exists("./temp/weather.bz2")) 
  {
    download.file(dataURL,"./temp/weather.bz2")
  }
  
  # check if reducedStormData variable already exists
  dataSet <- NULL
  if(file.exists("./temp/weathercache.RData"))
  {
    #load("./temp/weathercache.RData")
    #print(paste("Dataset loaded from cache. Rows:  ", nrow(dataSet)))
    #don't return dataset if it cache exists
  }
  else
  {
    dataSet <- read.csv("weather.bz2")
    print(paste("Dataset loaded from start. Rows:  ", nrow(dataSet)))
  }
  dataSet
}
dataSet <- readData()
```

### Transformation 

Dataset takes too long to process. Recude the size and keep only recent, more accurate years of observations. 

Columns to keep

 * **EVTYPE** for event analysis
 * **FATALITIES** and **INJURIES** for health damage
 * **-DMG-** columns for economic damage


Alphabetical characters used to signify magnitude include

 * **K** for thousands
 * **M** for millions
 * **B** for billions

If additional precision is available, it may be provided in the narrative part of the entry

```{r}
reduceSize <- function()
{
  if(file.exists("./temp/weathercache.RData"))
  {
    load("./temp/weathercache.RData")
    print(paste("slimSet loaded from cache. Rows:  ", nrow(slimSet)))
    return(slimSet)
  }
  dataSet$Date = as.Date(dataSet$BGN_DATE, format = "%m/%d/%Y")
  tempDS <- subset(dataSet,
                   Date > as.Date("01/01/2000",format = "%m/%d/%Y") &
                     EVTYPE != "?" & 
                     (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | PROPDMG > 0)
  )
  
  keepCols <- c("EVTYPE", "FATALITIES", "INJURIES", 
                "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP","Date")
  
  slimSet <- tempDS[,keepCols]
  
  
  slimSet$PropertyDamage <- mapply(fixAmount, as.numeric(slimSet$PROPDMG), slimSet$PROPDMGEXP)
  slimSet$CropDamage <- mapply(fixAmount, as.numeric(slimSet$CROPDMG), slimSet$CROPDMGEXP)
  save(slimSet, file="./temp/weathercache.RData")
  slimSet
}

fixAmount <- function(tempAmt, powerInd)
{
  amount <- NULL
  if (powerInd == "" | powerInd == " ")
  {
    amount <- tempAmt
  }
  else if (powerInd == "K")
  {
    amount <- tempAmt * 10^3
  }
  else if (powerInd == "M")
  {
    amount <- tempAmt * 10^6
  }
  else if (powerInd == "B")
  {
    amount <- tempAmt * 10^9
  }
  else 
  {
    print(paste(c("Unexpected power: ",powerInd)))
  }
  amount
}

slimSet <- reduceSize()
```

## Results

Get top 20 weather events that threaten public health by fatalities and injuries

```{r}
analyzeHealth <- function()
{
  eventFatalities <- aggregate(FATALITIES ~ EVTYPE, slimSet, sum)
  eventInjuries <- aggregate(INJURIES ~ EVTYPE, slimSet, sum)
  merged <- arrange(merge(eventFatalities, eventInjuries, by.x = "EVTYPE"), desc(FATALITIES))
  print(merged[1:20,])
  melted <- melt(merged[1:20,], c("EVTYPE"))
  
  
  g <- ggplot(melted, aes(x = reorder(factor(EVTYPE),-value), value)) +
    geom_bar(stat="identity", aes(fill = variable)) +
    xlab("Event type") +
    ylab("Health damage") +
    ggtitle('Extreme weather effects on public health year 2000+') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  print(g)
}
analyzeHealth()
```

Get top 20 weather events that damage property and crops

```{r}
analyzeEconomy <- function()
{
  propDamage <- aggregate(PropertyDamage ~ EVTYPE, slimSet, sum)
  cropDamage <- aggregate(CropDamage ~ EVTYPE, slimSet, sum)
  mergedEco <- arrange(merge(propDamage, cropDamage, by.x = "EVTYPE"), desc(PropertyDamage))
  print(mergedEco[1:20,])
  meltedEco <- melt(mergedEco[1:20,], c("EVTYPE"))
  
  g <- ggplot(meltedEco, aes(x = reorder(factor(EVTYPE),-value), value)) +
    geom_bar(stat="identity", aes(fill = variable)) +
    #geom_text(aes(label = round(value/10^9,1), y = value), size = 3) +
    xlab("Event type") +
    ylab("Economic damage (billions)") +
    ggtitle('Extreme weather damage to economy year 2000+') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  print(g)
}
analyzeEconomy()
```

## Conclusions
#### Public health analysis:
###### Tornadoes and excessive heat are the most damaging weather events with respect
###### to the fatalities and injuries among population.
#### Economy damage analysis:
###### Floods are by far the most devastating events to property and crops, followed by hurricanes and storm surges. The crop damage pales in comparrisson to personal and public property from any top weather event.